---
title: "STA426 - Week 11 - High-dimensional cytometry data analysis"
author: "Lukas M. Weber"
date: "11/26/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

See slides:

- Single-cell data
- High-dimensional cytometry
- Bioconductor
- CyTOF workflow and CATALYST vignette



# Install Bioconductor packages

See separate R script.



# Demo: workflow based on CyTOF workflow and CATALYST vignette


## Background

See text in Sections 1 (Introduction) and 2 (Data description) in CyTOF workflow.


## 1. Example data

Note: the example data used in the CATALYST vignette is a subset of the full BCR-XL
dataset from the CyTOF workflow.

```{r}
library(CATALYST)

# load example data
data(PBMC_fs, PBMC_panel, PBMC_md)

PBMC_fs
length(PBMC_fs)
PBMC_fs[[1]]
head(exprs(PBMC_fs[[1]]))

PBMC_panel

PBMC_md
```


## 2. Data organization: The daFrame class

The 'daFrame' class is used to structure the data for the differential analysis workflow.
This collapses the multiple tables of data (one table per sample) from the raw data into a
single table.

```{r}
# construct daFrame
daf <- daFrame(PBMC_fs, PBMC_panel, PBMC_md)

library(SummarizedExperiment)

daf
dim(daf)
head(exprs(daf))
rowData(daf)
colData(daf)
```


## 3. Diagnostic plots

3.1. plotCounts: Number of cells measured per sample

```{r}
n_cells(daf)

plotCounts(daf, color_by = "condition")
```


3.2. plotMDS: Multi-dimensional scaling plot

```{r}
plotMDS(daf, color_by = "condition")
```


3.3. plotExprHeatmap: Heatmap of (scaled) median marker expressions

```{r}
plotExprHeatmap(daf, color_by = "condition")
```


## 4.  Clustering

4.1. cluster: FlowSOM clustering & ConsensusClusterPlus metaclustering

```{r}
# specify markers to use for clustering
lineage_markers <- c("CD3", "CD45", "CD4", "CD20", "CD33", "CD123", "CD14", "IgM", "HLA_DR", "CD7")

# perform clustering
daf <- cluster(daf, cols_to_use = lineage_markers)

rowData(daf)
colData(daf)
```


Note: Skip Sections 4.2 (Delta area plot) and 4.3 (mergeClusters: Manual cluster merging)
for this demo.


4.4. plotClusterHeatmap: Heatmap of (meta)clustering results

Note: using default values for arguments 'k' and 'm'.

```{r}
# median pS6 expression by sample as 2nd heatmap
plotClusterHeatmap(daf, hm2 = "pS6")
```

```{r}
# population frequencies by sample as 2nd heatmap
plotClusterHeatmap(daf, hm2 = "abundances", draw_freqs = TRUE, cluster_anno = FALSE)
```


4.5. plotAbundances: Relative population abundances

Note: using default value for argument 'k'.

```{r}
plotAbundances(daf, by = "sample_id", group = "condition")
```

```{r}
plotAbundances(daf, by = "cluster_id", group = "condition")
```


## 5. Dimensionality reduction with tSNE

```{r}
daf <- tSNE(daf, cols_to_use = lineage_markers, 
            n = 200, verbose = FALSE, seed = 42)
```


5.1. plotSNE: t-SNE visualization

```{r}
# color by pS6 expression and split by condition
plotSNE(daf, color_by = "pS6", facet = "condition")
```


Note: using default value for argument 'color_by'.

```{r}
# color by metaclusters and split by sample ID
plotSNE(daf, facet = "sample_id")
```


## 6. Differental testing with 'diffcyt'

```{r}
library(diffcyt)

# create model formula
formula <- createFormula(PBMC_md, cols_fixed = 3, cols_random = 4)
formula

# create contrast
contrast <- createContrast(c(0, 1))
contrast

# test for differential abundance (DA) of clusters
res_DA <- diffcyt(daf, formula = formula, contrast = contrast, 
                  analysis_type = "DA", method_DA = "diffcyt-DA-GLMM", clustering_to_use = "meta20")

# test for differential states (DS) within clusters
res_DS <- diffcyt(daf, formula = formula, contrast = contrast, 
                  analysis_type = "DS", method_DS = "diffcyt-DS-LMM", clustering_to_use = "meta20")
```


6.1. plotDiffHeatmap: Heatmap of differential testing results

```{r}
# DA test results
plotDiffHeatmap(daf, res_DA, all = TRUE, normalize = TRUE)
```


```{r}
# DS test results
plotDiffHeatmap(daf, res_DS, top_n = 20, normalize = TRUE)
```



# Exercise

Perform a "null comparison" using code and data from the CATALYST vignette and/or CyTOF workflow.

–  e.g. 4 vs. 4 comparison of "Reference" samples only (instead of comparing stimulated vs. reference samples)
–  note: the CATALYST vignette contains a subset of the full dataset; if you need more samples, try downloading the full data using code from the CyTOF workflow
–  re-run the analysis by modifying inputs to the plotting functions and differential testing functions
–  plot and interpret the results
–  what do the results show? what did you expect?



