---
title: "Exercise 4"
author: "Hubert Rehrauer"
date: "07 10 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis

Do an exploratory data analysis of a matrix of expression values. Load the data and display:

* distributions: *boxplot*, *density*, *limma::plotDensities*
* normalization: *limma::normalizeQuantiles*
* clustering: *hclust*
* heatmap: use the package *pheatmap*
* correlation matrix: *cor* and *image*
* reduced dimensionality representation: *cmdscale* and *prcomp*


## Data Import
```{r }
anno = read.table("SampleAnnotation.txt", as.is=TRUE, sep="\t", quote="",
                  row.names=1, header=TRUE)
x = read.table("expressiondata.txt", as.is=TRUE, sep="\t", quote="", row.names=1, header=TRUE, check.names = FALSE)
x = as.matrix(x)
```

## Define samples and colors and phenotype
```{r}
samples = rownames(anno)
colors = rainbow(nrow(anno))
isNorm = anno$TissueType == "norm"
isSick = anno$TissueType == "sick"
isAcute = anno$TissueType == "acute"
```

## Notes

* When clustering the samples, please make sure that you cluster the columns, i.e. the distance matrix should be computed for the columns.
* Since the expression values span a large dynamic range, do a log-transform of the values before plotting
* For the heatmap use only the 500 genes with the highest variance across the samples

